---
title: "Reproducable Research Proj 1"
author: "Elizabeth Ellis"
date: "August 11, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading in Data

First, our data needs to be loaded into R. The data can be found in the github repository in the activity.zip file. The data will be saved in the R object 'data'.

```{r get data}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```

Next, we need to clean the data up. The dates need to be converted to the proper date formats. Weekdays are then calculated and added to the dataset 

```{r data cleanup}
activity$date <- as.POSIXct(activity$date, "%Y-%m-%d", tz = "EST")
weekday <- weekdays(activity$date)
activity <- cbind(activity,weekday)
```

Finally, let's load in the necessary R packages!

```{r}
require(ggplot2)
```


## What is the mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
```{r total steps}
activity_total_steps <- with(activity, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
names(activity_total_steps) <- c("date", "steps")
```

2. Create a histogram of the total number of steps taken per day
```{r histogram}
hist(activity_total_steps$steps, main = "Total Number of Steps Taken per Day", xlab = "Total Steps per Day", col = "pink", ylim = c(0,20), breaks = seq(0,25000, by=2500))
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r mean and median}
mean1 <- mean(activity_total_steps$steps)
median1 <- median(activity_total_steps$steps)
paste("The mean number of steps taken per a day is", mean1)
paste("The median number of steps taken per a day is", median1)
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r time series data}
average_daily_activity <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)
names(average_daily_activity) <- c("interval", "mean")
```
```{r time series chart}
plot(average_daily_activity$interval, average_daily_activity$mean, type = "l", col="orchid2", lwd = 2, xlab="Interval", ylab="Average Number of Steps", main="Average Number of Steps per Intervals")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r max steps}
maxint <- average_daily_activity[which.max(average_daily_activity$mean), ]$interval
paste("Interval number", maxint, "has the maximum number of steps.")
```

## Imputing Missing Values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missing values total}
misssteps <- sum(is.na(activity$steps))
paste("The number of missing values is", misssteps)
```

2. Devise a strategy for filling in all of the missing values in the dataset. (Replace missing values with mean score)

```{r missing values calc}
imputed_steps <- average_daily_activity$mean[match(activity$interval, average_daily_activity$interval)]
```

3. Create a new dataset with the missing values replaced by the mean

```{r new dataset}
activity_imputed <- transform(activity, steps = ifelse(is.na(activity$steps), yes = imputed_steps, no = activity$steps))
total_steps_imputed <- aggregate(steps ~ date, activity_imputed, sum)
names(total_steps_imputed) <- c("date", "daily_steps")
```

4. Make a histogram of the total number of steps taken each day

```{r new histogram}
hist(total_steps_imputed$daily_steps, main = "Total Number of Steps Taken per Day", xlab = "Total Steps per Day", col = "lightblue1", ylim = c(0,30), breaks = seq(0,25000, by=2500))
```

5. Calculate the new mean and median

```{r new mean and median}
mean2 <- mean(total_steps_imputed$daily_steps)
median2 <- median(total_steps_imputed$daily_steps)
paste("The new mean number of steps taken per a day is", mean2)
paste("The new median number of steps taken per a day is", median2)
```

6. How are the histogram, mean, and median different from the original dataset?

    - The central bin on the histogram is larger and the mean and median are both higher
    - The mean and median are now the same number. 

##Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r weekend factor}
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))
activity$datetype <- sapply(activity$date, function(x) {
        if (weekdays(x) == "Saturday" | weekdays(x) == "Sunday") 
                {y <- "Weekend"} else 
                {y <- "Weekday"}
                y
        })
```

2. Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r time series}
activity_by_date <- with(activity, aggregate(steps~interval + datetype, FUN = mean, na.rm = TRUE))
plot<- ggplot(activity_by_date, aes(x = interval , y = steps, color = datetype)) +
       geom_line() +
       labs(title = "Average Daily Steps by Type of Date", x = "Interval", y = "Average Number of Steps") +
       facet_wrap(~datetype, ncol = 1, nrow=2)
print(plot)
```

